---
title: Nesting with tidyr
author: Alex Zajichek
date: '2023-03-23'
slug: nesting-with-tidyr
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2023-03-23T06:10:16-05:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

_2/2/2023_

I really love the elegance of the [`nest`](https://tidyr.tidyverse.org/reference/nest.html) functionality with the [`tidyr`](https://tidyr.tidyverse.org/) package. It really allows you to abstract the meaning of a data frame to not just contain rectangular data with scalars, but rather a generalization that has rectangular data of _objects_. The most intriguing part of it to me is the way we can continue to use typical [`join`](https://dplyr.tidyverse.org/reference/mutate-joins.html) operations even with complex objects in some of the columns, which makes it so smooth and intuitive to do complex data operations.

```{r}
# Load packages
library(tidyverse)
```

For example, lets say we have a dataset.

```{r}
dat <- cheese::heart_disease
dat
```

And we want to compute age percentiles by sex for those who do and don't have heart disease.

```{r}

# Nest data frames within sex, heart disease
nested1 <-
  dat %>%
  group_by(Sex, HeartDisease) %>%
  nest()
nested1
```

We can see that there is now a separate dataset available within each combination of sex and heart disease status in the form of a `list` column.

```{r}
# Get the empirical cumulative density function for age
nested2 <-
  nested1 %>%
  mutate(
    ecdf_col = data %>% map(~ecdf(.x$Age))
  )
nested2
```

We then apply `list` operations as we normally would. In this case, we use `purrr::map` to create an empirical cumulative density function for age within each group. The result is then just a `list` of `ecdf` functions.

```{r}
# Make an age grid
age_grid <-
  dat %>%
  select(Sex, HeartDisease) %>%
  distinct() %>%
  inner_join(
    y = tibble(Age = c(40, 50, 60, 70)),
    by = character()
  )
age_grid
```

We then made an age grid for each sex/heart disease combination to evaluate the percentiles of each age in the respective groups. Now, we can compute the percentiles by joining to get the `ecdf` for the respective group, and plugging in each age into the function.

```{r}
age_grid %>% 
  
  # Join to get the ecdf for the group
  inner_join(
    y = nested2 %>% select(-data),
    by = c("Sex", "HeartDisease")
  ) %>%
  
  # Compute the percentile for the given age
  mutate(
    Percentile = map2(.x = ecdf_col, .y = as.list(Age), .f = ~.x(.y)) %>% flatten_dbl()
  )
```

We can see, for example, that a 60 year old male is at the $86^{th}$ percentile for those without heart disease, but at the $72^{nd}$ for those who due, suggesting that the age distribution tends to be higher in patients with heart disease.