---
title: 'Hospital Readmissions: Building a Filterable Map with leaflet and crosstalk'
subtitle: Series on runtime-free content delivery
author: Alex Zajichek
date: '2022-06-18'
slug: filterable-maps
categories: []
tags: []
authors: []
lastmod: '2022-06-18T07:49:36-05:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
summary: So you run `R` locally but want to be able to deliver engaging, interactive analytic content to your stakeholders. Is there a way to do that?
---

_**In Progress**_

Want to make this map widget? Read On! `r emo::ji("backhand index pointing down")``r emo::ji("backhand index pointing down")``r emo::ji("backhand index pointing down")`

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE, comment = FALSE, results = 'hide'}

### Load packages
require(tidyverse)
require(readxl)
require(tigris)
require(sf)
require(leaflet)
require(crosstalk)

### 1. Import CMS data

## i. FY22 final rule (indicates penalties for each hospital)

# Set location of zip file (publicly available on CMS website)
hrrp_zip <- "https://www.cms.gov/files/zip/fy2022-hrrp-final-rule-supplemental-file.zip"

# Create a temporary file (Credits: rpubs.com/otienodominic/398952)
temp_file <- tempfile()

# Download into the temporary file
download.file(hrrp_zip, temp_file)

# Name of file needed within zip
hrrp_file <- "FY2022_Final_Rule_Supplemental_File.xlsx"

# Unzip, and place the file in the current working directory
unzip(temp_file, hrrp_file, exdir = ".")

# Import the data file into a data frame
hrrp_results <-
  read_xlsx(
    path = hrrp_file,
    sheet = "FR FY 2022",
    skip = 3,
    na = c("", " ", ".", "-", "NA", "N/A")
  )

# Delete the downloaded file
file.remove(hrrp_file)
unlink(temp_file)

## ii. Hospital Information

# Set path to file (from CMS)
hosp_file <- "https://data.cms.gov/provider-data/sites/default/files/resources/092256becd267d9eeccf73bf7d16c46b_1641873938/Hospital_General_Information.csv"

# Import the file
hosp_info <-
  read_csv(
    file = hosp_file
  )

### 2. Obtain zip code centroids

# Get set of zip codes for Wisconsin
zips_wi <- zctas(year = 2010, state = "WI")

# Gather centroids of corrdinates for each zip
zips_wi_centroids <-
  zips_wi %>%
  
  # Get the centroid
  st_centroid() %>%
  
  # Pluck the coordinates
  st_coordinates() %>%
  
  # Make a tibble
  as_tibble() %>%
  
  # Add identifying column
  add_column(
    Zip = zips_wi$ZCTA5CE10
  ) %>%
  
  # Rename columns
  rename(
    lon = X,
    lat = Y
  )

### 3. Build an analysis dataset
map_data <-
  hosp_info %>%
  
  # Filter to Wisconsin hospitals
  filter(
    State == "WI"
  ) %>%
  
  # Keep a few pieces of information
  select(
    FacilityID = `Facility ID`,
    FacilityName = `Facility Name`,
    Zip = `ZIP Code`
  ) %>%
  
  # Join to get the centroid for the hospital's zip code
  inner_join(
    y = zips_wi_centroids,
    by = "Zip"
  ) %>%
  
  # Add random jitter to coordinates
  mutate(
    across(
      c(
        lat,
        lon
      ),
      jitter,
      amount = 0.05
    )
  ) %>%
  
  # Join to get HRRP program results
  inner_join(
    y = 
      hrrp_results %>%
      
      # Make cleaner levels for penalty indicators
      mutate(
        across(
          starts_with("Penalty indicator"),
          ~
            case_when(
              .x == "Y" ~ "Yes",
              .x == "N" ~ "No",
              TRUE ~ NA_character_
            )
        )
      ) %>%
      
      # Keep a few columns
      select(
        FacilityID = `Hospital CCN`,
        PeerGroup = `Peer group assignment`,
        DualProportion = `Dual proportion`,
        Penalty = `Payment Reduction Percentage`,
        starts_with("Penalty indicator")
      ) %>%
      
      # Remove the prefix from the cohort columns
      rename_with(
        ~str_remove(.x, "^Penalty indicator for "),
        starts_with("Penalty indicator")
      ),
    by = "FacilityID"
  ) %>% 
  
  # Join to get list of penalized cohorts (for labeling)
  left_join(
    y = 
      hrrp_results %>%
      
      # Send down the rows
      pivot_longer(
        cols = starts_with("Penalty indicator"),
        names_prefix = "Penalty indicator for "
      ) %>%
      
      # Filter to penalty cohorts
      filter(
        value == "Y"
      ) %>%
      
      # For each hospital
      group_by(`Hospital CCN`) %>%
      
      # Concatenate the list of penalty cohorts
      summarise(
        PenalizedCohortCount = n(),
        PenalizedCohorts = paste(sort(unique(name)), collapse = ", "),
        .groups = "drop"
      ) %>%
      
      # Rename the column
      rename(
        FacilityID = `Hospital CCN`
      ),
    by = "FacilityID"
  ) %>%
  
  # Fill in non-penalized hospitals
  mutate(
    PenalizedCohortCount = coalesce(PenalizedCohortCount, 0),
    PenalizedCohorts = coalesce(PenalizedCohorts, "")
  ) 

### 4. Map data components

# State outline
state_outline <-
  maps::map(
    database = "state",
    regions = "wisconsin",
    fill = TRUE,
    plot = FALSE
  )

# County outlines
county_outlines <- 
  counties(cb = TRUE) %>%
  filter(
    STATE_NAME == "Wisconsin"
  )

### 4. Build map

## i. State outline
my_map <- 
  leaflet() %>%
  
  # Add geographic tiles
  addTiles() %>%
  
  # Add WI state outline
  addPolygons(
    data = state_outline,
    fillColor = "gray",
    stroke = FALSE
  ) 

## ii. County outline
my_map <-
  my_map %>%
  
  # Add county outlines
  addPolygons(
    data = county_outlines,
    color = "black",
    fillColor = "#ff59f7",
    weight = 1,
    opacity = .5,
    fillOpacity = .35,
    highlightOptions = 
      highlightOptions(
        color = "black",
        weight = 3,
        bringToFront = FALSE
      ),
    label = ~NAME
  )

```

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE, comment = FALSE}

## iii. Add hospitals

# Make the shared data set
sd <- SharedData$new(data = map_data)

# Make filters
bscols(
  widths = c(1, 11),
  "",
  htmltools::h3("Hospital Readmissions Reduction Program (HRRP) FY 2022 - WI Map", style = "font-style:italic;")
)
# Row 1:
bscols(
  widths = c(1, 10, 1),
  "",
  filter_slider(
    id = "penalty",
    label = "Payment Reduction",
    sharedData = sd,
    column = ~Penalty,
    step = 0.1,
    min = 0,
    max = 3,
    post = "%"
  ),
  ""
)

# Row 2-3:
bscols(
  widths = c(1, 11),
  "",
  "Cohort-specific penalty:"
)
bscols(
  widths = c(1, 3, 3, 3, 1),
  "",
  filter_select(
    id = "ami",
    label = "AMI",
    sharedData = sd,
    group = ~AMI
  ),
  filter_select(
    id = "cabg",
    label = "CABG",
    sharedData = sd,
    group = ~CABG
  ),
  filter_select(
    id = "copd",
    label = "COPD",
    sharedData = sd,
    group = ~COPD
  ),
  ""
)
bscols(
  widths = c(1, 3, 3, 3, 1),
  "",
  filter_select(
    id = "hf",
    label = "HF",
    sharedData = sd,
    group = ~HF
  ),
  filter_select(
    id = "pneu",
    label = "PNEU",
    sharedData = sd,
    group = ~pneumonia
  ),
  filter_select(
    id = "tja",
    label = "THA/TKA",
    sharedData = sd,
    group = ~`THA/TKA`
  ),
  ""
)

# Create a color pallete
pal <- 
  colorNumeric(
    palette = "RdYlGn",
    domain = -1*seq(0, 3, .1)
  )

# Add shared data to the map
my_map %>%
  
  addCircleMarkers(
    data = sd, # A dataset containing coordinates for each zip code (or anything)
    lng = ~lon, # The coordinates of that entitity
    lat = ~lat,
    label = ~paste0(FacilityName, " (click for info)"),
    popup = 
      ~paste0(
        "Hospital: ", FacilityName, 
        "<br>Zip Code: ", Zip,
        "<br>Peer Group: ", PeerGroup,
        "<br>Dual-Eligibility: ", round(DualProportion*100, 2), "%",
        "<br>Penalties: ", PenalizedCohorts,
        "<br>Payment Reduction: ", Penalty, "%"
      ),
    color = ~pal(-1*Penalty),
    fillOpacity = .75
  )

```

<br>
<br>

So you run `R` locally but want to be able to deliver engaging, interactive analytic content to your stakeholders. Is there a way to do that? Lately, I've spent a lot of time learning and exploring tools to find the answer to that question. Luckily, the answer is _yes_, because there are many amazing packages that have been developed to help with this, including [`reactable`](https://glin.github.io/reactable/), [`plotly`](https://plotly.com/r/), [`DT`](https://rstudio.github.io/DT/), [`leaflet`](https://rstudio.github.io/leaflet/), and [`crosstalk`](https://rstudio.github.io/crosstalk/), among others. This article focuses on the latter two, and how we can build map widgets that are changeable by the user through filters, all in a self-contained HTML document.

# Table of Contents

* [The Use-Case: Hospital Readmissions](#usecase)
  + [Creating the dataset](#usecasedataset)
* [Building the Map](#mapoverview)
  1. [Create the base layer](#baselayer)
* [Conclusion](#conclusion)
* [Code Appendix](#codeappendix)

# Defining a Use-Case: Hospital Readmissions {#usecase}

A _readmission_ occurs when a patient returns to a hospital for acute care within 30 calendar days of previously being discharged. Each year the [_Centers for Medicare & Medicaid Services (CMS)_](https://www.cms.gov/) [penalize hospitals](https://www.cms.gov/Medicare/Medicare-Fee-for-Service-Payment/AcuteInpatientPPS/Readmissions-Reduction-Program) across the United States that have _excess_ readmissions in a select set of high-risk patient populations. This is determined by

1. Adjusting for the hospital's specific patient population acuity by a collection of clinical risk factors
2. Comparing the expected readmission rate of _those_ patients had they been treated at _your_ hospital versus the _national average_ hospital. The difference between those rates yield the penalty amount (with some additional steps and caveats).

For the stats gurus, it is just a mixed-effects logistic regression model that uses the hospital as a random-intercept term, and many clinical factors as fixed effects. The hospital's specific estimated effect is plugged into the regression to compute the _predicted_ readmission rate. You can read more about the methodology [here](https://qualitynet.cms.gov/inpatient/measures/readmission/methodology).

The goal of this exercise is to build a map widget to identify and explore participating Wisconsin hospitals who were penalized in the _FY 2022 Hospital Readmissions Reduction Program (HRRP)_.

## Creating the dataset {#usecasedataset}

# Building the Map {#mapoverview}

## 1. Create the base layer {#baselayer}