---
title: Removing higher-order aggregation text in {reactable}
author: Alex Zajichek
date: '2023-06-29'
slug: quickly-removing-higher-order-aggregation-text-in-reactable
categories: []
tags: ['rstats', 'javascript']
subtitle: 'A quick & dirty workaround'
summary: ''
authors: []
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Often when I'm building a [`reactable`](https://glin.github.io/reactable/) I have potentially many aggregation levels but only want to display data in a lower subset of those. Here is a quick way to eliminate text using cell styling. 

First let's load some packages and look at an example data set.

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(reactable)
dat <- cheese::heart_disease
dat
```

Suppose we are interested in these outcomes: 

1. The rate of heart disease
2. The rate of patients with a maximum heart rate > 150 

And we want to summarize those, along with a few other patient characteristics, within a handful of important clinical factors. We might pivot our data into a long format to start:

```{r}
pivot_dat <-
  dat |> 
  
  # Convert to character types
  mutate(
    across(
      c(ChestPain, ExerciseInducedAngina, BloodSugar),
      as.character
    ),
    MaximumHR =
      case_when(
        MaximumHR > 150 ~ "Yes",
        TRUE ~ "No"
      )
  ) |>
  
  # Send "important clinical factors" down the rows
  pivot_longer(
    cols = c(ChestPain, ExerciseInducedAngina, BloodSugar),
    names_to = "ClinicalFactor",
    values_to = "Level"
  ) |> 
  
  # Send outcomes down the rows
  pivot_longer(
    cols = c(HeartDisease, MaximumHR),
    names_to = "Outcome",
    values_to = "HasOutcome"
  )
pivot_dat
```

We can then calculate some summary statistics. Our full patient population is replicated for each `ClinicalFactor` and `Outcome`, so we should account for that.

```{r}
summary_dat <-
  pivot_dat |> 
  
  # Summarize characteristics
  summarize(
    Patients = n(),
    OutcomeRate = mean(HasOutcome == "Yes"),
    Age = mean(Age),
    Female = mean(Sex == "Female"),
    .by = 
      c(
        ClinicalFactor,
        Level,
        Outcome,
        HasOutcome
      )
  ) |> 
  
  # Add counts within groups
  mutate(
    PercentOfPatients = Patients / sum(Patients),
    .by = c(ClinicalFactor, Outcome)
  ) |> 
  
  # Make clean levels/labels
  mutate(
    ClinicalFactor = 
      case_when(
        ClinicalFactor == "BloodSugar" ~ "Blood sugar > 120 mg/dl?",
        ClinicalFactor == "ChestPain" ~ "Chest pain type",
        ClinicalFactor == "ExerciseInducedAngina" ~ "Exercise-induced angina?"
      ),
    Level = 
      case_when(
        Level == "TRUE" ~ "Yes",
        Level == "FALSE" ~ "No",
        TRUE ~ Level
      ),
    Outcome = 
      case_when(
        Outcome == "HeartDisease" ~ "Heart disease",
        Outcome == "MaximumHR" ~ "Maximum HR > 150"
      )
  ) |>
  
  # Rearrange
  relocate(PercentOfPatients, .after = Patients) |>
  relocate(Outcome, .before = everything()) |>
  arrange(
    Outcome,
    ClinicalFactor,
    Level,
    HasOutcome
  )
summary_dat
```

Now we're ready to build the table (note the use of the [`zildge::rectbl_agg_wtd()`](https://github.com/zajichek/zildge/blob/main/R/rectbl_agg_wtd.R) function to compute the weighted average, see [my other post](https://www.zajichekstats.com/post/reactable-javascript/) for more information about this).

```{r}
reactable(
  data = summary_dat,
  groupBy = c("Outcome", "ClinicalFactor", "Level"),
  columns = 
    list(
      ClinicalFactor = colDef(name = "Clinical factor"),
      HasOutcome = colDef(name = "Has Outcome?"),
      Patients = colDef(name = "Count", aggregate = "sum", align = "center"),
      PercentOfPatients = colDef(name = "Percent", aggregate = "sum", align = "center", format = colFormat(digits = 2, percent = TRUE)),
      OutcomeRate = colDef(name = "Outcome rate (%)", aggregate = zildge::rectbl_agg_wtd("Patients"), align = "center", format = colFormat(digits = 2, percent = TRUE)),
      Age = colDef(name = "Avg. age (years)", aggregate = zildge::rectbl_agg_wtd("Patients"), align = "center", format = colFormat(digits = 2)),
      Female = colDef(name = "Female (%)", aggregate = zildge::rectbl_agg_wtd("Patients"), align = "center", format = colFormat(digits = 2, percent = TRUE))
    ),
  columnGroups = 
    list(
      colGroup(
        name = "Patients",
        columns = c("Patients", "PercentOfPatients")
      )
    ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  resizable = TRUE,
  theme = reactablefmtr::sandstone()
) |>
  reactablefmtr::add_source("Use arrows to expand table", font_size = 12, font_style = "italic")
```

<br>

There are a few problems with this table:

1. The redundancy of the top-level (and second-level) summary statistics is unappealing
2. The top-level aggregation is just flat-out wrong because we've duplicated counts over the data set

Maybe _1_ is not a huge deal, but I've probably already established somewhere that there are `r nrow(dat)` patients in the data set and the overall rate of heart disease is `r round(100*mean(dat$HeartDisease == "Yes"),1)`%. You still may choose to keep it as-is in this case as it's not totally harmful. However, _2_ needs to be addressed as the data is misleading and not really interpretable. __So how can we remove the results from (at least) the top-level while maintaining the lower-level aggregation?__

We can define an `R` function that provides JavaScript [custom styling](https://glin.github.io/reactable/articles/conditional-styling.html) to a cell.

```{r}
remove_text <-
  function(col, threshold) {
    JS(
      paste0(
        "function(rowInfo) {
          if(rowInfo.row['", col, "'] > ", threshold, ") {
            return {fontSize:0}
          }
        }"
      )
    )
  }
```

All this function does is reduce the font size to zero for any cell where the value of `col` in the respective row is greater than `threshold`. Thus, since we are over-counting the number of patients in the top-level, we can just apply this function to rows in which the count exceeds `r nrow(dat)`, which is the patient count in our data set. We provide this function to the `style` argument of `colDef()`.

```{r}
reactable(
  data = summary_dat,
  groupBy = c("Outcome", "ClinicalFactor", "Level"),
  columns = 
    list(
      ClinicalFactor = colDef(name = "Clinical factor"),
      HasOutcome = colDef(name = "Has Outcome?"),
      Patients = colDef(style = remove_text("Patients", nrow(dat)), name = "Count", aggregate = "sum", align = "center"),
      PercentOfPatients = colDef(style = remove_text("Patients", nrow(dat)), name = "Percent", aggregate = "sum", align = "center", format = colFormat(digits = 2, percent = TRUE)),
      OutcomeRate = colDef(style = remove_text("Patients", nrow(dat)), name = "Outcome rate (%)", aggregate = zildge::rectbl_agg_wtd("Patients"), align = "center", format = colFormat(digits = 2, percent = TRUE)),
      Age = colDef(style = remove_text("Patients", nrow(dat)), name = "Avg. age (years)", aggregate = zildge::rectbl_agg_wtd("Patients"), align = "center", format = colFormat(digits = 2)),
      Female = colDef(style = remove_text("Patients", nrow(dat)), name = "Female (%)", aggregate = zildge::rectbl_agg_wtd("Patients"), align = "center", format = colFormat(digits = 2, percent = TRUE))
    ),
  columnGroups = 
    list(
      colGroup(
        name = "Patients",
        columns = c("Patients", "PercentOfPatients")
      )
    ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  resizable = TRUE,
  theme = reactablefmtr::sandstone()
) |>
  reactablefmtr::add_source("Use arrows to expand table", font_size = 12, font_style = "italic")
```

<br>

If we want to remove text from the second level of aggregation, we can just adjust the `threshold`.

```{r}
reactable(
  data = summary_dat,
  groupBy = c("Outcome", "ClinicalFactor", "Level"),
  columns = 
    list(
      ClinicalFactor = colDef(name = "Clinical factor"),
      HasOutcome = colDef(name = "Has Outcome?"),
      Patients = colDef(style = remove_text("Patients", nrow(dat) - 1), name = "Count", aggregate = "sum", align = "center"),
      PercentOfPatients = colDef(style = remove_text("Patients", nrow(dat) - 1), name = "Percent", aggregate = "sum", align = "center", format = colFormat(digits = 2, percent = TRUE)),
      OutcomeRate = colDef(style = remove_text("Patients", nrow(dat) - 1), name = "Outcome rate (%)", aggregate = zildge::rectbl_agg_wtd("Patients"), align = "center", format = colFormat(digits = 2, percent = TRUE)),
      Age = colDef(style = remove_text("Patients", nrow(dat) - 1), name = "Avg. age (years)", aggregate = zildge::rectbl_agg_wtd("Patients"), align = "center", format = colFormat(digits = 2)),
      Female = colDef(style = remove_text("Patients", nrow(dat) - 1), name = "Female (%)", aggregate = zildge::rectbl_agg_wtd("Patients"), align = "center", format = colFormat(digits = 2, percent = TRUE))
    ),
  columnGroups = 
    list(
      colGroup(
        name = "Patients",
        columns = c("Patients", "PercentOfPatients")
      )
    ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  resizable = TRUE,
  theme = reactablefmtr::sandstone()
) |>
  reactablefmtr::add_source("Use arrows to expand table", font_size = 12, font_style = "italic")
```

<br>

And that's it.

I'll acknowledge that I didn't exhaust all options for this task and that there are _definitely_ better ways to do this with more directed intention by accessing various properties of the table. Obviously this only works if you have a column in the table that only increases during the aggregation and does so to predictable amounts. But it did the trick here.
